---
title: 'De novo MPNST Drug Study: RNAseq in mouse DROP Veh outlier March 26, MYO batch adjust'
author: "Michael S. Chimenti"
date: "03.29.20"
output:
  html_document: default
  html_notebook: default
---

```{r setup, include=FALSE}
require("knitr")
knitr::opts_chunk$set(echo = TRUE)
opts_knit$set(root.dir = "~/collab_proj/quelle/project_rnaseq_mouse_MPNST_feb2021/")

```

**This is the third revision to this analysis to incorporate a myosin/troponin expression batch effect in the analysis.**


### Samples were prepared according to: 

The library prep used the TruSeq mRNA stranded kit and then they ran paired ends at 2x150 bp.

### Upstream processing 

Samples were processed with `bcbio v 1.2.4` using the `mm10` genome build from Ensembl. `hisat2` and `salmon` were used for alignment. 

### Example methods 

Here is a methods paragraph that captures what we did: 

**Analysis was conducted by the Bioinformatics Division of the Iowa Institute of Human Genetics (IIHG).  Reads were processed with the ‘bcbio-nextgen.py’ open-source informatics pipeline developed primarily at Harvard Chan Bioinformatics (v.1.0.8) [32] running on the Argon HPC resource at the University of Iowa.  This pipeline includes ‘best practices’ approaches for read quality control, read alignment and quantitation.  The ‘bcbio-nextgen.py’ pipeline was run in “RNA-seq” mode with the ‘mm10’ key as the selected genome build (internally referencing Ensembl GRCm38.p6 v94, GENCODE M19).  The pipeline aligned reads to the GRCm38 genome using the splice-aware, ultra-rapid hisat2 aligner (2.1.0) [34,35] and concurrently quantified reads to the GENCODE M19 transcriptome using the ‘salmon’ (0.9.1) aligner [36].  Qualimap (2.2.2) [33], a computational tool that examines hisat2 BAM alignment files, was used to examine the read data for quality control. We observed > X% of RNA-seq reads uniquely mapping to the reference and >X% of reads in exonic regions.  Sequence quality scores passed basic checks, and sequence duplication rates were within acceptable parameters.  Salmon-derived transcript quantifications (TPM) were imported and summarized to estimated counts at the gene level using tximport (1.12.3) in Rstudio [37], as described in the best-practices DESeq2 vignette (https://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html).  Genes with fewer than 5 estimated counts across all samples were pre-filtered from downstream analysis, as per recommended procedure.  Differential gene expression analysis was conducted with DESeq2(1.24.0) on estimated gene-level counts [38].  An FDR of 5% and X < abs(logFC) < 10 was set as a cutoff for differential expression (DEGs).**



<details>
  <summary>Code: Package Imports </summary>
  <p>
  
```{r, imports, warnings=FALSE, message=FALSE, eval=TRUE}
#negative binomial GLM and related
library('DESeq2')
library('calibrate')
library('tximport')
library('readr')
library('rhdf5')
#Exploratory analysis
library('tidyverse')
library('pcaExplorer')
library('ggrepel')
library('kableExtra')
#pathway and gene clusters
library('DEGreport')



```

 </p>
</details>


```{r, function_defs, eval=TRUE, echo = FALSE}

## Volcano Plot
do_vol_plot <- function(df, sig=0.05, fc=0.5, size = 4){
  
  df_sig<- df %>% filter(padj < sig)
  
  ## genes with labels -- signficant,decreased expression and start with IGF
  
  df_label<- df %>% 
              filter(padj < sig, abs(log2FoldChange) > fc)
              #mutate(short_id = stringr::str_split(gene_id, "000000") %>% map_chr(.,2))
  
  
  ## plot
  
  p <- ggplot(df, aes(log2FoldChange, -log10(padj)))  +
      geom_point(size=0.8, color="black", alpha=.8) + 
      geom_point(size=0.8, data=df_sig, aes(log2FoldChange, -log10(padj)), colour="red") +
       geom_text_repel(size= size, 
                    colour="black",
                    segment.size=0.1, 
                    nudge_x=0.06, 
                    nudge_y=0.06, 
                    data=df_label,
                    aes(log2FoldChange, -log10(padj), label=gene_name), 
                    max.iter= 200, 
                    point.padding = 0.15, 
                    segment.alpha = 1, 
                    box.padding=.15,
                    min.segment.length = unit(0.15, 'lines'),size=2.5) +
    theme(
      axis.text.x = element_text(size=12),
      axis.text.y = element_text(size=12),
      axis.title.x = element_text(size=14, margin = margin(t = 10, r = 0, b = 10, l = 0)),
      axis.title.y = element_text(size=14, margin = margin(t = 0, r = 10, b = 0, l = 10)),
      plot.margin =unit(c(.5,.5,.5,.5),"cm"),
      plot.title = element_text(size = 11)
    )
  
  return (p)
}

#function to quickly get sorted results 
get_res <- function(dds, meta_col, cond1, cond2, anno, pval = 0.05){
  res <- results(dds, contrast = c(meta_col,cond1,cond2))
  res <- na.omit(res)
  res_sig <- res[res$padj < pval,]
  res_ord <- res_sig[order(res_sig$padj),]
  return(res_ord)
}


get_res_iPG <- function(dds, meta_col, cond1, cond2, anno, pval = 0.05){
  res <- results(dds, contrast = c(meta_col,cond1,cond2))
  res <- na.omit(res)
  res <- res[abs(res$log2FoldChange) < 10, ]
  res$gene_id <- row.names(res)
  res <- left_join(as_tibble(res), as_tibble(anno), by = "gene_id")
  write.table(res, file = paste0("res_for_iPath_", cond1, "_", cond2, ".txt"), sep = '\t', 
              row.names=FALSE)}

anno <- readRDS("~/Desktop/anno_mm10.rds")
anno <- na.omit(anno)
```


 </p>
</details>

### Importing the data and converting to counts for DESeq2

The metadata table looks like this: 
```{r, import_meta}
samples <- read.table("samples.csv", sep=',', header=TRUE)
head(samples, 10)
```

```{r, sample_import, echo=FALSE}
rownames(samples) <- samples$sample
samples$sample <- as.character(samples$sample)

files <- file.path(getwd(), samples$sample, 'salmon', 'quant.sf')
names(files) <- samples$sample
```

Now, we import 'salmon' quantification tables, using the `tx2gene` object to correlate transcripts to genes: 
```{r, tximport, cache=TRUE}
tx2gene <- read.csv(file.path(getwd(), "tx2gene.csv"), header = FALSE, as.is = c(1:2)) 
txi <- tximport(files, type="salmon", tx2gene=tx2gene)
```

### DESeq2 analysis 

First, we'll create a DESeq2 object 
```{r, make_DEseq, warning = FALSE}
ddsTxi <- DESeqDataSetFromTximport(txi,
                                   colData = samples,
                                   design = ~ sex + cond)

ddsTxi <- ddsTxi[ rowSums(counts(ddsTxi)) > 5, ]
ddsTxi <- DESeq(ddsTxi)
```

### Large batch effect along PC2 is myosin expression 

```{r, new_covariate, eval = TRUE}
tab <- counts(ddsTxi) %>% as_tibble(rownames = "gene_id") %>% filter(gene_id %in% c("ENSMUSG00000055775", "ENSMUSG00000061086", "ENSMUSG00000009214", "ENSMUSG00000091898", "ENSMUSG00000091898", "ENSMUSG00000026414"))

tab2 <- tab %>% reshape2::melt() %>% group_by(variable) %>% mutate(ave_exp = mean(value))
tab3 <- tab2 %>% select(c(variable,ave_exp)) %>% unique()
tab3$myo_batch <- cut(tab3$ave_exp, breaks = c(0,500,10000), label = c("low_myo_exprs","high_myo_exprs"))


colnames(tab3) <- c("sample","ave_exp","myo")
colData(ddsTxi)$myo <- tab3$myo
vstTxi <- vst(ddsTxi)
p <- pcaExplorer::pcaplot(vstTxi, intgroup = c("myo"), ellipse = FALSE, text_labels = TRUE, point_size = 2, ntop=500, pcX = 1, pcY = 2)
p <- p + ggtitle("Myosin, troponin, actin, GPCR, SRY, interleukin33 levels")
p

```

### Drop the outlier "VehUR8FFT" AND account for "myo" batch 

I will add "myo" batch factor to the design to account for this obvious variation: 
```{r, drop_out}
ddsTxi_drop <- ddsTxi[,!ddsTxi$sample == "VehUR8FFT"]
design(ddsTxi_drop) <- ~ sex + myo + cond
ddsTxi_drop <- DESeq(ddsTxi_drop)
```


```{r, plot_combo_heatmap, dpi = 300, fig.width = 8, fig.height=8}
library(pheatmap)
library(RColorBrewer)

## Read in genes for heatmap
genes_of_int <- read.csv("heatmap_genes.csv", header = FALSE)
colnames(genes_of_int) <- "gene_name"
genes_of_int <- left_join(genes_of_int, anno, by = "gene_name")

# drop the duplicate anno for ccl19
genes_of_int <- genes_of_int[-18,]

## Add go term factor 
go_term <-data.frame(GO_term = c(rep("Humoral_immune_resp",20), 
                                 rep("Cell_cycle", 17)))
                                 
genes_of_int <- cbind(genes_of_int, go_term)


## Get variance stabilized counts for plot, create matrix for pheatmap
vstTxi_drop <- vst(ddsTxi_drop)
vst_tab <- as.data.frame(assay(vstTxi_drop))
counts_mat <- as.matrix(vst_tab[genes_of_int$gene_id,])



#cond <- data.frame(sample = rep(c("NK_TCell_Dep", "NK_Naive", "NK_PBMC"), 4), row.names = annot$sample)  ##

## Setup quantile breaks
quantile_breaks <- function(xs, n = 10) {
  breaks <- quantile(xs, probs = seq(0, 1, length.out = n))
  breaks[!duplicated(breaks)]
}

mat_breaks <- quantile_breaks(as.matrix(vst_tab), n = 13)  ## quantile breaks


#mat_colors <- list(cond = brewer.pal(3, "Set1"))

annot <- colData(vstTxi_drop)
anno_df <- as.data.frame(annot, row.names = colnames(vstTxi_drop))
colnames(anno_df) <- c("sample","sex","tumor_type","myo")
labels <- genes_of_int$gene_name

row_anno <- as.data.frame(genes_of_int$GO_term, row.names = genes_of_int$gene_id)
colnames(row_anno) <- "GO_term"

p <- pheatmap(counts_mat[,c(1,4,5,2,6,7,8,3,9,10,11)], 
         labels_row = labels,
         #breaks = mat_breaks,
         annotation_col = anno_df[,c("tumor_type","sex")],
         annotation_row = row_anno,
         #annotation_colors = mat_colors,
         cluster_rows = FALSE,
         cluster_cols = FALSE,
         show_rownames = TRUE,
         show_colnames = TRUE,
         color = colorRampPalette(rev(brewer.pal(n=10,name = "RdBu")))(15),
         fontsize_row = 8,
         scale = "row"
         )


ggsave(file = "heatmap_vehDROP_blu_red_ZscoreRow_ordered.pdf", plot=p, dev = "pdf")

```


```{r, PCA_explorer_1, echo=FALSE, dpi = 200}

vstTxi_drop <- vst(ddsTxi_drop, blind=FALSE)
colnames(vstTxi_drop) <- vstTxi_drop$sample

p <- pcaExplorer::pcaplot(vstTxi_drop, intgroup = c("cond"), ellipse = FALSE, text_labels = TRUE, point_size = 2, ntop=500, pcX = 1, pcY = 2)
p <- p + ggtitle("Vehicle clusters together with outlier dropped")
p

p <- pcaExplorer::pcaplot(vstTxi_drop, intgroup = c("myo"), ellipse = FALSE, text_labels = TRUE, point_size = 2, ntop=500, pcX = 1, pcY = 2)
p <- p + ggtitle("Myo and tropo levels still cluster samples")
p
```

#### Unfortunately, the control samples are confounded with the high "myo" batch.  

### DEG calculation
**NOTE:I redid this section filtering OUT any DEG with abs(log2FC) > 10.  These extremely large logFC DEGs are likely artifacts.  They should be ignored.**


```{r, get_res, echo = FALSE}

## here we define getting an annotated results tibble 
get_res_tib_anno <- function(dds, meta_col, cond1, cond2, anno, pval = 0.05){
  res <- results(dds, contrast = c(meta_col,cond1,cond2))
  res <- na.omit(res)
  res_sig <- res[res$padj < pval & abs(res$log2FoldChange) < 10,]
  res_ord <- res_sig[order(res_sig$padj),]
  res_ord$gene_id <- row.names(res_ord)
  res_ord <- left_join(as_tibble(res_ord), anno, by = "gene_id")
  return(res_ord)
}
```

### Combo resistant vs vehicle, Volcano Plot 

```{r, res1, dpi = 200, warning = FALSE, message = FALSE}
res_combo_res <- get_res_tib_anno(ddsTxi_drop, "cond", "combo_res", "veh", pval = 0.1, anno = anno)
p1 <- do_vol_plot(res_combo_res, sig = 0.05, fc = 1.0, size = 2)
p1 <- p1 + ggtitle("268 DEGs Res vs Veh (red = DEGS at FDR of 0.05)")
p1
```

### There are 268 DEGs between combo_resistant and "vehicle" at a FDR = 0.1

### Top 12 DE genes Combo Resistant vs Vehicle Count Plots 
```{r, gene_plot1, dpi = 200, fig.width=9, warning = FALSE, message = FALSE}
res <- results(ddsTxi_drop, contrast = c("cond","combo_res","veh"))
res <- na.omit(res)
res <- res[abs(res$log2FoldChange) < 10,]
DEGreport::degPlot(ddsTxi_drop, xs = "cond", res = res, n = 12, group = "sex", 
                   xsLab = "Treatment", ysLab = "raw counts")

```

### Top 20 DEGs table between combo_resistant and vehicle

```{r, table1}
my_cols <- c("log2FoldChange","pvalue","padj","gene_id","gene_name","description")
head(res_combo_res[,my_cols], 20) %>% kable() %>% kable_styling()

```


### Combo sensitive vs vehicle; Volcano Plot 

```{r, res2, dpi = 200, warning = FALSE, message=FALSE}
res_combo_sen <- get_res_tib_anno(ddsTxi_drop, "cond", "combo_sen", "veh", pval = 0.1, anno = anno)

p1 <- do_vol_plot(res_combo_sen, sig = 0.005, fc = 1.0, size = 3)
p1 <- p1 + ggtitle("There are 600 DEGs Sens vs Veh (Red = DEGs at FDR = 0.005)")
p1
```



### Top 12  DE genes Combo Sensitive vs Vehicle Count Plots 
```{r, gene_plot2, dpi = 200, fig.width = 9, warning = FALSE, message = FALSE}
res <- results(ddsTxi_drop, contrast = c("cond","combo_sen","veh"))
res <- na.omit(res)
res <- res[abs(res$log2FoldChange) < 10,]
DEGreport::degPlot(ddsTxi_drop, xs = "cond", res = res, n = 12, group = "sex", 
                   xsLab = "Treatment", ysLab = "raw counts")
```

### Top 20 DEGs table between combo_sensitive and vehicle

```{r, table2}
head(res_combo_sen[,my_cols], 20) %>% kable() %>% kable_styling()

```


### Combo Sensitive vs Combo Resistant 

#### Positive log FC is UP in combo sensitive 

```{r, res3, dpi = 200, warning = FALSE, message = FALSE}
res_sens_vs_resist <- get_res_tib_anno(ddsTxi_drop, "cond", "combo_sen", "combo_res", pval = 0.1, anno = anno)
p1 <- do_vol_plot(res_sens_vs_resist, sig = 0.01, fc = 1.0, size = 3)
p1 <- p1 + ggtitle("300 DEGs between combo_sens and combo_resist (red = FDR 0.01)")
p1
```



### Top 12 DEGS plot Combo Sens vs Combo Res 
```{r, gene_plot3, dpi = 200, fig.width = 9, warning = FALSE, message = FALSE}
res <- results(ddsTxi_drop, contrast = c("cond","combo_sen","combo_res"))
res <- na.omit(res)
res <- res[abs(res$log2FoldChange) < 10,]
DEGreport::degPlot(ddsTxi_drop, xs = "cond", res = res, n = 12, group = "sex", 
                   xsLab = "Treatment", ysLab = "raw counts")
```


### Top 20 DEGS Table Combo Sens vs Combo Resistant

**We can see that Immunoglobin Light Chains dominate the DEGs between the two tumor types**

```{r, table3}
head(res_sens_vs_resist[,my_cols], 20) %>% kable() %>% kable_styling()
```


### VENN analysis 

How do the these gene lists compare to each other?  

```{r, venn, warning=FALSE, message=FALSE }
library(Vennerable)
v <- Vennerable::Venn(list(combo_res_DEGs = res_combo_res$gene_id, combo_sen_DEGs = res_combo_sen$gene_id, sens_vs_res_DEGs = res_sens_vs_resist$gene_id))
plot(v, doWeights = TRUE)

```

### How do the adjustments (outlier drop, myo batch effect) affect the outcome? 


#### Dropping outlier expands the pool of DEGs (adds 639) for Sens vs Veh

```{r, venn2, warning=FALSE, message=FALSE }
library(Vennerable)

## Gene list no outlier drop, no myo batch correct 
res_sens_nodrop_nomyo <- get_res_tib_anno(ddsTxi, "cond", "combo_sen", "veh", pval = 0.1, anno = anno)

## Drop outlier, no batch correct 
ddsTxi_drop2 <- ddsTxi[,!ddsTxi$sample == "VehUR8FFT"]
ddsTxi_drop2 <- DESeq(ddsTxi_drop2)

## Results from outlier drop, no batch correct 
res_sens_drop_nomyo <- get_res_tib_anno(ddsTxi_drop2, "cond", "combo_sen", "veh", pval = 0.1, anno = anno)
## Results from outlier drop AND batch correct 
res_sens_drop_myo <- get_res_tib_anno(ddsTxi_drop, "cond", "combo_sen", "veh", pval = 0.1, anno = anno)

v1 <- Vennerable::Venn(list(sens_DEGs_noDrop_noMyo = res_sens_nodrop_nomyo$gene_id, 
                           sens_DEGs_Drop_noMyo = res_sens_drop_nomyo$gene_id))
plot(v1, doWeights = TRUE)

```



#### "Myo" batch modeling shrinks DEG pool substantially

**241 DEGs are "discovered" when modeling for Myo batch**

```{r, venn3, warning=FALSE, message=FALSE }

v2 <- Vennerable::Venn(list(sens_DEGs_Drop_noMyo = res_sens_drop_nomyo$gene_id, 
                           sens_DEGs_Drop_Myo = res_sens_drop_myo$gene_id))
plot(v2, doWeights = TRUE)
```

#### All three approaches compared

```{r, venn4, warning=FALSE, message=FALSE }
v3 <- Vennerable::Venn(list(sens_DEGs_noDrop_noMyo = res_sens_nodrop_nomyo$gene_id, 
                           sens_DEGs_Drop_noMyo = res_sens_drop_nomyo$gene_id, 
                           sens_DEGs_Drop_Myo = res_sens_drop_myo$gene_id))
plot(v3, doWeights = TRUE)

```



```{r, write_gene_lists, echo=FALSE}

readr::write_csv(res_combo_res[,my_cols], "DEGs_combo_resistant_vs_vehicle_FDR_10percent_VEHUR8_OUT_DROP_MYO_BATCH.csv")
readr::write_csv(res_combo_sen[,my_cols], "DEGs_combo_sensitive_vs_vehicle_FDR_10percent_VEHUR8_OUT_DROP_MYO_BATCH.csv")
readr::write_csv(res_sens_vs_resist[,my_cols], "DEGS_combo_sensitive_vs_combo_resistant_FDR_10percent_VEHUR8_OUT_DROP_MYO_BATCH.csv")

#get_res_iPG(dds = ddsTxi_drop, meta_col = "cond", cond1 = "combo_sen", cond2 = "veh", anno = anno)
#get_res_iPG(dds = ddsTxi_drop, meta_col = "cond", cond1 = "combo_res", cond2 = "veh", anno = anno)
#get_res_iPG(dds = ddsTxi_drop, meta_col = "cond", cond1 = "combo_sen", cond2 = "combo_res", anno = anno)

```





```{r, sessionInfo}
sessionInfo()
```